---
title: "Correcting allantois experiments"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/01-allantois-correction.Rmd/",
  cache.lazy = FALSE,
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r}
pkg <- c("BiocParallel", "SingleCellExperiment", "scater", "scran", "uwot", "Rtsne", "batchelor", "patchwork", "EnsDb.Mmusculus.v79", "org.Mm.eg.db", "msigdbr")

lib <- lapply(pkg, library, character.only = TRUE)
```

## Data

### Allantois

Read processed experiment data:

```{r}
sce1 <- readRDS("data/allantois-wt.rds")

sce2 <- readRDS("data/allantois-kdr.rds")
```

Extract original experiment data:

```{r}
alt1 <- altExp(sce1, "original", withColData = TRUE)

alt2 <- altExp(sce2, "original", withColData = TRUE)
```

Make cell names unique across experiments:

```{r}
colnames(alt1) <- paste0("allantois-wt-", seq_len(ncol(alt1)))

colnames(alt2) <- paste0("allantois-kdr-", seq_len(ncol(alt2)))
```

Make feature names compatible across experiments:

```{r}
rownames(alt1) <- rowData(alt1)$gene_id

rownames(alt2) <- rowData(alt2)$gene_id
```

Make altExp compatible across experiments:

```{r}
altExp(alt1, "ERCC") <- NULL

altExp(alt2, "ERCC") <- NULL
```

## Integration

### Variance modelling

Collect variance modelling statistics:

```{r}
dec1 <- metadata(alt1)$modelGeneVarWithSpikes

dec2 <- metadata(alt2)$modelGeneVarWithSpikes
```

Make features names compatible across experiments:

```{r}
rownames(dec1) <- rowData(alt1)$gene_id

rownames(dec2) <- rowData(alt2)$gene_id
```

Identify common feature names across experiments:

```{r}
ids1 <- rownames(dec1)

ids2 <- rownames(dec2)

ids <- intersect(ids1, ids2)
```

Combine variance modelling results from both studies:

```{r}
dec1 <- dec1[ids, ]

dec2 <- dec2[ids, ]

dec <- combineVar(dec1, dec2)

rownames(dec) <- ids
```

### Feature selection

Filter Xist gene from list of features:

```{r}
hvg <- rownames(dec)

lgl <- hvg == "ENSMUSG00000086503" # Xist

hvg <- hvg[!lgl]
```

Filter Y chromosome genes from list of features:

```{r}
ids <- keys(EnsDb.Mmusculus.v79, keytype = "GENEID", filter = SeqNameFilter("Y"))

lgl <- hvg %in% ids

hvg <- hvg[!lgl]
```

Filter MT chromosome genes from list of features:

```{r}
ids <- keys(EnsDb.Mmusculus.v79, keytype = "GENEID", filter = SeqNameFilter("MT"))

lgl <- hvg %in% ids

hvg <- hvg[!lgl]
```

Filter cell-cycle genes from list of features:

```{r}
org <- select(org.Mm.eg.db, keys = "GO:0007049", keytype = "GOALL", column = "ENSEMBL")

lgl <- hvg %in% org$ENSEMBL

hvg <- hvg[!lgl]
```

Filter ribosomal genes from list of features:

```{r}
ids <- select(EnsDb.Mmusculus.v79, keys = keys(EnsDb.Mmusculus.v79), columns = c("GENEID", "GENENAME"), keytype = "GENEID")

sig <- msigdbr(species = "Mus musculus", category = "C2")

sig <- subset(sig, gs_name == "KEGG_RIBOSOME")

org <- select(org.Mm.eg.db, keys = sig$gene_symbol, keytype = "SYMBOL", column = "ENSEMBL")

lgl <- hvg %in% org$ENSEMBL

hvg <- hvg[!lgl]
```

Select features for batchelor integration:

```{r eval = FALSE}
hvg <- getTopHVGs(dec[hvg, ], prop = 0.1)
```

Define a set of markers to use for correction:
  
```{r eval = TRUE}
res1 <- pairwiseWilcox(logcounts(alt1), alt1$cluster, direction = "up")

sig1 <- getTopMarkers(res1$statistics, res1$pairs, n = 10)

res2 <- pairwiseWilcox(logcounts(alt2), alt2$cluster, direction = "up")

sig2 <- getTopMarkers(res2$statistics, res2$pairs, n = 10)

ids <- unique(unlist(c(unlist(sig1), unlist(sig2))))

hvg <- ids[ids %in% hvg]
```

### Batch normalization

Identify common feature names across experiments:

```{r}
ids1 <- rownames(alt1)

ids2 <- rownames(alt2)

ids <- intersect(ids1, ids2)
```

Subset experiments to common universe of features:

```{r}
alt1 <- alt1[ids, ]

alt2 <- alt2[ids, ]
```

Per-batch scaling normalization:

```{r}
out <- multiBatchNorm("allantois-wt" = alt1, "allantois-kdr" = alt2, norm.args = list(use_altexps = FALSE))

alt1 <- out[["allantois-wt"]]

alt2 <- out[["allantois-kdr"]]
```

### Batch combination

Combine experiments with no correction:

```{r}
all <- correctExperiments("allantois-wt" = alt1, "allantois-kdr" = alt2, PARAM = NoCorrectParam())
```

Calculate PCA reduced dimensions:

```{r}
set.seed(1101011110)

all <- runPCA(all, subset_row = hvg)
```

Create PCA plot coloured by batch:

```{r}
plotPCA(all, colour_by = "batch")
```

Create PCA plot coloured by celltype:

```{r}
plotPCA(all, colour_by = "celltype")
```

Calculate UMAP reduced dimensions:

```{r}
set.seed(1101011110)

all <- runUMAP(all, dimred = "PCA")
```

Create UMAP plot coloured by batch:

```{r}
plotUMAP(all, colour_by = "batch")
```

Create UMAP plot coloured by celltype:

```{r}
plotUMAP(all, colour_by = "celltype")
```

### Batch Correction

Perform fast mutual nearest neighbors correction:

```{r}
set.seed(1010111011)

mnn <- correctExperiments(all, batch = all$batch, subset.row = hvg, correct.all = TRUE)
```

Check proportion of lost variance to ensure genuine biological heterogeneity is not being removed:

```{r}
metadata(mnn)$merge.info$lost.var
```

## Exploration

### MNN embedding

Create MNN plot coloured by batch:

```{r}
plotReducedDim(mnn, "corrected", colour_by = "batch")
```

Create MNN plot coloured by celltype:

```{r}
plotReducedDim(mnn, "corrected", colour_by = "celltype")
```

### UMAP embedding

Calculate UMAP using different parameters:

```{r eval = FALSE}
num <- c(5, 15, 30, 50, 100)

dst <- c(0, 0.01, 0.05, 0.1, 0.5, 1)

arg <- expand.grid(n_neighbors = num, min_dist = dst)

funUMAP <- function(x, dimred, n_neighbors, min_dist) {
  
  set.seed(1011101101)
  
  runUMAP(x, dimred = dimred, n_neighbors = n_neighbors, min_dist = min_dist)

}

run <- mapply(
  FUN = funUMAP,
  n_neighbors = arg$n_neighbors,
  min_dist = arg$min_dist,
  MoreArgs = list(x = mnn, dimred = "corrected"),
  SIMPLIFY = FALSE
)
```

Select optimal UMAP parameters:

```{r}
set.seed(1011101101)

mnn <- runUMAP(mnn, dimred = "corrected", n_neighbors = 30, min_dist = 0.5)
```

Create UMAP plot coloured by batch:

```{r}
plotUMAP(mnn, colour_by = "batch")
```

Create UMAP plot coloured by celltype:

```{r}
plotUMAP(mnn, colour_by = "celltype")
```

## Summary

### Output

Write experiment object to a file:

```{r}
saveRDS(mnn, file = "data/01-allantois-correction.rds")
```

### Session

Print version information about R, the OS and attached or loaded packages:

```{r}
sessionInfo()
```
