---
title: "Trajectory analysis (revisited)"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/07-trajectory-analysis-revisited.Rmd/",
  cache.lazy = FALSE,
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r}
pkg <- c("BiocNeighbors", "EnsDb.Mmusculus.v79", "MouseGastrulationData", "RColorBrewer", "SingleCellExperiment", "cluster", "scater", "scran", "slingshot", "ggbeeswarm", "pheatmap", "ggbeeswarm")

lib <- lapply(pkg, library, character.only = TRUE)
```

Create colour palettes:

```{r}
EmbryoGenotypeColours <- c("wt" = "#7fc97f", "kdr" = "#f0027f")

EmbryoSourceColours = c("allantois" = "#386cb0", "atlas" = "#beaed4")
```

Create nicer labels:

```{r}
EmbryoGenotypeLabels <- c("wt" = "WT", "kdr" = "KO")

EmbryoSourceLabels <- c("allantois" = "Allantois", "atlas" = "Atlas")

EmbryoNeighborLabels <- c("TRUE" = "Neighbor", "FALSE" = "Non-Neighbor")
```

## Processing

Read experiment data:

```{r}
mnn <- readRDS("data/02-atlas-integration.rds")
```

Get feature name for each identifier:

```{r}
rowData(mnn)$gene_id <- rownames(mnn)

rowData(mnn)$gene_name <- mapIds(EnsDb.Mmusculus.v79, keys = rownames(mnn), column = "GENENAME", keytype = "GENEID")

rowData(mnn)$gene_name[rowData(mnn)$gene_name == ""] <- NA
```

Make feature names unique:

```{r}
rownames(mnn) <- uniquifyFeatureNames(rowData(mnn)$gene_id, rowData(mnn)$gene_name)
```

## Nearest neighbors

Query k-nearest neighbors to each allantois cell:

```{r}
set.seed(503179283)

dim <- reducedDim(mnn, "corrected")

sbj <- dim[mnn$source == "atlas", ]

qry <- dim[mnn$source == "allantois", ]

knn <- queryKNN(sbj, qry, k = 1)

idx <- unique(as.vector(knn$index))

ids <- rownames(sbj)[idx]

lgl <- colnames(mnn) %in% ids

mnn$neighbor <- ifelse(lgl, TRUE, FALSE)
```

Create plotting data for exploration:

```{r}
dim <- reducedDim(mnn, "UMAP")

colnames(dim) <- c("X", "Y")

dat <- as.data.frame(mnn@colData)

tbl <- cbind(dim, dat)

bgd <- as.data.frame(dim)
```

Plot UMAP projection coloured by celltype / facetted by neighbor and source:

```{r fig.width = 14, fig.height = 9}
ggplot(tbl, aes(X, Y)) + 
  geom_point(data = bgd, colour = "gainsboro", size = 0.1) + 
  geom_point(mapping = aes(colour = celltype), size = 0.1) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(ncol = 1, override.aes = list(size = 2))) + 
  facet_grid(neighbor ~ source, labeller = labeller(neighbor = EmbryoNeighborLabels, source = EmbryoSourceLabels)) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.title = element_blank())
```

Plot UMAP projection coloured by genotype / facetted by neighbor and source:

```{r fig.width = 14, fig.height = 9}
ggplot(tbl, aes(X, Y)) + 
  geom_point(data = bgd, colour = "gainsboro", size = 0.1) + 
  geom_point(mapping = aes(colour = genotype), size = 0.1) + 
  scale_colour_manual(name = "Genotype", values = EmbryoGenotypeColours, labels = EmbryoGenotypeLabels) + 
  guides(colour = guide_legend(ncol = 1, override.aes = list(size = 2))) + 
  facet_grid(neighbor ~ source, labeller = labeller(neighbor = EmbryoNeighborLabels, source = EmbryoSourceLabels)) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.title = element_blank())
```

## Diffusion Map

Select wild-type allantois and k-nearest neighbor cells:

```{r}
id1 <- mnn$source == "allantois"

id2 <- mnn$neighbor == TRUE

sel <- id1 | id2

sce <- mnn[, sel]
```

Create a diffusion map of cells from the batch-corrected PCA matrix:

```{r}
set.seed(312898858)

sce <- runDiffusionMap(sce, dimred = "corrected")
```

Identify 'optimal' number of clusters to build a minimum spanning tree in slingshot:

```{r}
set.seed(695895185)

dim <- reducedDim(sce, "DiffusionMap")

fit <- clusGap(dim, kmeans, K.max = 20)

num <- maxSE(fit$Tab[, "gap"], fit$Tab[, "SE.sim"])
```

Cluster the selected cell types to achieve a more granular trajectory:

```{r}
set.seed(231362263)

dim <- reducedDim(sce, "DiffusionMap")

fit <- kmeans(dim, centers = 8)

sce$cluster <- factor(fit$cluster)
```

Setup colour palette for clusters:

```{r}
EmbryoClusterColours <- brewer.pal(nlevels(sce$cluster), name = "Set3")

names(EmbryoClusterColours) <- levels(sce$cluster)
```

Create plotting data for exploration:

```{r fig.width = 7.0, fig.height = 4.5}
dim <- reducedDim(sce, "DiffusionMap")

colnames(dim) <- c("X", "Y")

dat <- as.data.frame(sce@colData)

tbl <- cbind(dim, dat)

bgd <- as.data.frame(dim)
```

Plot DiffusionMap projection coloured by cluster / facetted by genotype and source:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(X, Y)) + 
  geom_point(data = bgd, colour = "gainsboro", size = 2) + 
  geom_point(mapping = aes(colour = cluster), size = 2) + 
  scale_colour_manual(name = "Cluster", values = EmbryoClusterColours) + 
  guides(colour = guide_legend(ncol = 1, override.aes = list(size = 2))) + 
  facet_grid(genotype ~ source, labeller = labeller(genotype = EmbryoGenotypeLabels, source = EmbryoSourceLabels)) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.title = element_blank())
```

Plot DiffusionMap projection coloured by celltype / facetted by genotype and source:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(X, Y)) + 
  geom_point(data = bgd, colour = "gainsboro", size = 2) + 
  geom_point(mapping = aes(colour = celltype), size = 2) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  guides(colour = guide_legend(ncol = 1, override.aes = list(size = 2))) + 
  facet_grid(genotype ~ source, labeller = labeller(genotype = EmbryoGenotypeLabels, source = EmbryoSourceLabels)) +
  theme_bw() +
  theme(aspect.ratio = 1, axis.title = element_blank())
```

## Slingshot

Perform lineage inference with Slingshot:

```{r}
sce <- slingshot(
  data = sce,
  clusterLabels = sce$cluster,
  reducedDim = "DiffusionMap",
  start.clus = 5
)
```

Create SlingshotDataSet from SingleCellExperiment object:

```{r}
sds <- SlingshotDataSet(sce)
```

Create plotting data for exploration:

```{r}
dim <- reducedDim(sce, "DiffusionMap")

colnames(dim) <- c("X", "Y")

dat <- as.data.frame(sce@colData)

tbl <- cbind(dim, dat)

sc1 <- as.data.frame(sds@curves$curve1$s[sds@curves$curve1$ord, ])

colnames(sc1) <- c("X", "Y")

sc2 <- as.data.frame(sds@curves$curve2$s[sds@curves$curve2$ord, ])

colnames(sc2) <- c("X", "Y")

sc3 <- as.data.frame(sds@curves$curve3$s[sds@curves$curve3$ord, ])

colnames(sc3) <- c("X", "Y")
```

Plot DiffusionMap with lineage 1 coloured by pseudotime:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(X, Y)) + 
  geom_point(aes(fill = slingPseudotime_1), size = 3, shape = 21) + 
  scale_fill_viridis_c(name = "Pseudotime", option = "viridis") +  
  geom_path(data = sc1, aes(X, Y), size = 1.5, arrow = arrow()) + 
  labs(x = "DC1", y = "DC2") + 
  theme_bw() +
  theme(aspect.ratio = 1)
```

Plot DiffusionMap with lineage 2 coloured by pseudotime:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(X, Y)) + 
  geom_point(aes(fill = slingPseudotime_2), size = 3, shape = 21) + 
  scale_fill_viridis_c(name = "Pseudotime", option = "viridis") +  
  geom_path(data = sc2, aes(X, Y), size = 1.5, arrow = arrow()) + 
  labs(x = "DC1", y = "DC2") + 
  theme_bw() +
  theme(aspect.ratio = 1)
```

Plot DiffusionMap with lineage 3 coloured by pseudotime:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(X, Y)) + 
  geom_point(aes(fill = slingPseudotime_3), size = 3, shape = 21) + 
  scale_fill_viridis_c(name = "Pseudotime", option = "viridis") +  
  geom_path(data = sc3, aes(X, Y), size = 1.5, arrow = arrow()) + 
  labs(x = "DC1", y = "DC2") + 
  theme_bw() +
  theme(aspect.ratio = 1)
```



Plot DiffusionMap with lineages and coloured by cluster:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(X, Y)) + 
  geom_point(aes(fill = cluster), size = 3, shape = 21) + 
  scale_fill_manual(name = "Cluster", values = EmbryoClusterColours) + 
  geom_path(data = sc1, aes(X, Y), size = 1.5, arrow = arrow()) + 
  geom_path(data = sc2, aes(X, Y), size = 1.5, arrow = arrow()) + 
  geom_path(data = sc3, aes(X, Y), size = 1.5, arrow = arrow()) + 
  labs(x = "DC1", y = "DC2") + 
  theme_bw() +
  theme(aspect.ratio = 1)
```

Plot DiffusionMap with lineages and coloured by celltype:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(X, Y)) + 
  geom_point(aes(fill = celltype), size = 3, shape = 21) + 
  scale_fill_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  geom_path(data = sc1, aes(X, Y), size = 1.5, arrow = arrow()) + 
  geom_path(data = sc2, aes(X, Y), size = 1.5, arrow = arrow()) + 
  geom_path(data = sc3, aes(X, Y), size = 1.5, arrow = arrow()) + 
  labs(x = "DC1", y = "DC2") + 
  theme_bw() +
  theme(aspect.ratio = 1)
```

Plot DiffusionMap with lineages and coloured by genotype:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(X, Y)) + 
  geom_point(aes(fill = genotype), size = 3, shape = 21) + 
  scale_fill_manual(name = "Genotype", values = EmbryoGenotypeColours, labels = EmbryoGenotypeLabels) + 
  geom_path(data = sc1, aes(X, Y), size = 1.5, arrow = arrow()) + 
  geom_path(data = sc2, aes(X, Y), size = 1.5, arrow = arrow()) + 
  labs(x = "DC1", y = "DC2") + 
  theme_bw() +
  theme(aspect.ratio = 1)
```

Plot cells ordered by pseudotime 1 and coloured by celltype / facetted by source:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(slingPseudotime_1, genotype, colour = celltype)) + 
  geom_quasirandom(groupOnX = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  scale_y_discrete(labels = EmbryoGenotypeLabels) + 
  facet_wrap(~ source, labeller = labeller(source = EmbryoSourceLabels)) + 
  labs(title = "Lineage 1", x = "Slingshot pseudotime", y = "Genotype") + 
  theme_bw() +
  theme(aspect.ratio = 1)
```

Plot cells ordered by pseudotime 2 and coloured by celltype / facetted by source:

```{r fig.width = 7.0, fig.height = 4.5}
ggplot(tbl, aes(slingPseudotime_2, genotype, colour = celltype)) + 
  geom_quasirandom(groupOnX = FALSE) + 
  scale_colour_manual(name = "Celltype", values = EmbryoCelltypeColours) + 
  scale_y_discrete(labels = EmbryoGenotypeLabels) + 
  facet_wrap(~ source, labeller = labeller(source = EmbryoSourceLabels)) + 
  labs(title = "Lineage 2", x = "Slingshot pseudotime", y = "Genotype") + 
  theme_bw() +
  theme(aspect.ratio = 1, axis.title = element_blank())
```

## Pseudotime correlation

Create function to identify genes that are significantly correlated with pseudotime:

```{r}
correlateTime <- function(x, column, alpha = 0.01) {
  
  nan <- is.na(x[[column]])

  x <- x[, !nan]
  
  mat <- assay(x, "logcounts") # use logcounts

  mat <- as.matrix(mat)

  mat <- rbind(mat, pseudotime = x[[column]])

  idx <- list("pseudotime", rownames(mat))
  
  res <- correlatePairs(mat, block = x$batch, pairings = idx) # block on batch

  sig <- subset(res, FDR < alpha)
  
  return(sig)

}
```

Identify genes correlated with lineage 1 pseudotime:

```{r}
set.seed(19167957)

de1 <- correlateTime(sce[,  !is.na(sce$slingPseudotime_1)], column = "slingPseudotime_1", alpha = 0.05)

DT::datatable(as.data.frame(de1))
```

Identify genes correlated with lineage 2 pseudotime:

```{r}
set.seed(605782840)

de2 <- correlateTime(sce[,  !is.na(sce$slingPseudotime_2)], column = "slingPseudotime_2", alpha = 0.05)

DT::datatable(as.data.frame(de2))
```

Identify genes correlated with lineage 3 pseudotime:

```{r}
set.seed(605782840)

de3 <- correlateTime(sce[,  !is.na(sce$slingPseudotime_3)], column = "slingPseudotime_3", alpha = 0.05)

DT::datatable(as.data.frame(de3))
```

Save results table of significant genes for downstream analysis:

```{r}
metadata(sce)$correlateTime <- list(slingPseudotime_1 = de1, slingPseudotime_2 = de2, slingPseudotime_3 = de3)
```

Create function to plot heatmap of genes expression values ordered by pseudotime:

```{r}
plotTimeGenes <- function(x, features, columns, order_columns_by) {

  obj <- x[features, columns]

  mat <- t(scale(t(assay(obj, "reconstructed"))))

  mat <- mat[, order(obj[[order_columns_by]])]

  col <- colorRampPalette(rev(brewer.pal(5, name = "RdBu")))(100)

  lim <- max(abs(mat))
  
  brk <- seq(-5, 5, length.out = 101)
  
  ann <- list(
    
    col = data.frame(
      Celltype  = obj$celltype,
      Genotype  = obj$genotype,
      Source    = obj$source,
      Cluster   = obj$cluster,
      row.names = colnames(obj)
    ),
    
    colors = list(
      Celltype = EmbryoCelltypeColours[unique(obj$celltype)],
      Genotype = EmbryoGenotypeColours[unique(obj$genotype)],
      Source   = EmbryoSourceColours[unique(obj$source)],
      Cluster  = EmbryoClusterColours[unique(obj$cluster)]
    )
  )

  plt <- pheatmap(
    mat = mat,
    color = col, 
    breaks = brk,
    scale = "none",
    cluster_cols = FALSE,
    annotation_col = ann$col,
    annotation_colors = ann$colors,
    show_colnames = FALSE
  )
  
  return(plt)

}
```

Plot expression of 50 most significant genes correlated with lineage 1 pseudotime:

```{r fig.width = 14, fig.height = 9}
plotTimeGenes(sce, features = head(de1$gene2, n = 50), columns = complete.cases(sce$slingPseudotime_1), order_columns_by = "slingPseudotime_1")
```

Plot expression of 50 most significant genes correlated with lineage 2 pseudotime:

```{r fig.width = 14, fig.height = 9}
plotTimeGenes(sce, features = head(de2$gene2, n = 50), columns = complete.cases(sce$slingPseudotime_2), order_columns_by = "slingPseudotime_2")
```

Plot expression of 50 most significant genes correlated with lineage 3 pseudotime:

```{r fig.width = 14, fig.height = 9}
plotTimeGenes(sce, features = head(de3$gene2, n = 50), columns = complete.cases(sce$slingPseudotime_3), order_columns_by = "slingPseudotime_2")
```

## Summary

Save experiment object with DiffusionMap and Slingshot results:

```{r}
saveRDS(sce, file = "data/07-trajectory-analysis-revisited.rds")
```

Print session information:

```{r}
sessionInfo()
```
