---
title: "Trajectory modules"
author: "James Ashmore"
date: "2020-04-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/09-trajectory-modules.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load Bioconductor packages:

```{r}
library(limma)
library(org.Mm.eg.db)
library(scater)
library(scran)
```

Load CRAN packages:

```{r}
library(ggplot2)
library(reshape2)
```

Read experiment data:

```{r}
sce <- readRDS("data/04-differential-analysis.rds")
```

## Utils

Define function to rescale the expression data:

```{r}
rescale <- function(x) {
  
  x <- as.matrix(x)
	
  nsamples <- ncol(x)
  
  M <- rowMeans(x, na.rm = TRUE)
	
  DF <- nsamples - 1L
	
  isNA <- is.na(x)
	
  if (any(isNA)) {
    
    mode(isNA) <- "integer"
		
    DF <-  DF - rowSums(isNA)
		
    DF[DF == 0L] <- 1L
	
  }
	
  x <- x - M
	
  V <- rowSums(x^2L, na.rm = TRUE) / DF
	
  x <- x / sqrt(V + 0.01)

}
```

Define function to reorder experiment data:

```{r}
reorder <- function(x, subset, select) {
  
  row <- subset %in% rownames(x)
  
  col <- order(select)
  
  x[row, col]

}
```

Define wrapper for the goana function:

```{r}
calculateGO <- function(query, subject) {
  
  res <- goana(de = query, universe = subject, species = "Mm")
  
  res <- split(res, res$Ont)
  
  res <- lapply(res, function(x) x[order(x$P.DE, decreasing = FALSE), ])
  
  res <- lapply(res, function(x) { x$FDR <- p.adjust(x$P.DE, method = "fdr"); x} )
  
  res <- lapply(res, subset, FDR < 0.05)
  
  return(res)

}
```

Define function to visualise GO results:

```{r}
visualiseGO <- function(results, n = 10) {
  
  results <- head(results, n = n)
  
  if (nrow(results) == 0) { return(ggplot() + theme_void()) }
  
  results$Term <- tools::toTitleCase(results$Term)
  
  results$Ratio <- results$DE / results$N

  plt <- ggplot(results, aes(x = Ratio, y = stats::reorder(Term, Ratio), colour = -log10(FDR), size = DE)) + 
    geom_point() + 
    scale_colour_viridis_c() + 
    labs(x = "Ratio", size = "Count", colour = "-log10(FDR)") + 
    guides(colour = guide_colorbar(order = 1), size = guide_legend(order = 2)) +
    theme_bw() + 
    theme(axis.title.y = element_blank())
  
  return(plt)

}
```

Define wrapper for the kegga function:

```{r}
calculateKEGG <- function(query, subject) {
  
  res <- kegga(de = query, universe = subject, species = "Mm")

  res <- res[order(res$P.DE, decreasing = FALSE), ]

  res$FDR <- p.adjust(res$P.DE, method = "fdr")
  
  res <- sig <- subset(res, FDR < 0.05)
  
  return(res)

}
```

Define function to visualise KEGG results:

```{r}
visualiseKEGG <- function(results, n = 10) {
  
  results <- head(results, n = n)
  
  if (nrow(results) == 0) { return(ggplot() + theme_void()) }
  
  results$Pathway <- tools::toTitleCase(results$Pathway)
  
  results$Ratio <- results$DE / results$N

  plt <- ggplot(results, aes(x = Ratio, y = stats::reorder(Pathway, Ratio), colour = -log10(FDR), size = DE)) + 
    geom_point() + 
    scale_colour_viridis_c() + 
    labs(x = "Ratio", size = "Count", colour = "-log10(FDR)") + 
    guides(colour = guide_colorbar(order = 1), size = guide_legend(order = 2)) +
    theme_bw() + 
    theme(axis.title.y = element_blank())
  
  return(plt)

}
```

## Modules

Rescale the expression data prior to clustering:

```{r}
assay(sce, "rescaled") <- rescale(assay(sce, "reconstructed"))
```

Define a set of background universe genes:

```{r}
dec <- modelGeneVar(sce, block = sce$batch)

hvg <- getTopHVGs(dec, var.threshold = 0) # take all genes with biological variation

ens <- rowData(sce)$gene_id[match(hvg, rownames(sce))]

uni <- mapIds(org.Mm.eg.db, keys = ens, column = "ENTREZID", keytype = "ENSEMBL")
```

### Lineage 1

#### Clustering

Extract association test results from experiment metadata:

```{r}
res <- metadata(sce)$associationTest
```

Create data frame of association test results for slingPseudotime_1 lineage:

```{r}
sp1 <- data.frame(
  waldStat = res$slingPseudotime_1$waldStat_1, 
  df = res$slingPseudotime_1$df_1, 
  pvalue = res$slingPseudotime_1$pvalue_1, 
  fdr = p.adjust(res$slingPseudotime_1$pvalue_1, method = "fdr"),
  row.names = rownames(res$slingPseudotime_1)
)
```

Reorder experiment data by slingPseudotime_1 lineage:

```{r}
sc1 <- reorder(sce, subset = rownames(sp1), select = sce$slingPseudotime_1)

use <- !is.na(sc1$slingPseudotime_1)

sc1 <- sc1[, use]
```

Cluster expression patterns of significant genes in slingPseudotime_1 lineage:

```{r}
set.seed(515499208)

km1 <- kmeans(x = assay(sc1, "rescaled"), centers = 3)

sp1$cluster <- km1$cluster[rownames(sp1)]
```

Plot clustered expression patterns to see how they behave across the slingPseudotime_1 lineage:

```{r}
clust.centroid <- function(i, assay, cluster) {
  ind <- (cluster == i)
  colMeans(assay[ind, ])
}

lvl <- unique(km1$cluster)

df1 <- sapply(lvl, clust.centroid, assay(sc1, "rescaled"), km1$cluster)

df1 <- melt(df1)

colnames(df1) <- c("cell", "cluster", "exprs")

df1$pseudotime <- sc1$slingPseudotime_1[df1$cell]

ggplot(df1, aes(pseudotime, exprs, group = cluster, colour = as.factor(cluster))) + 
  geom_point() + 
  scale_colour_brewer(name = "Cluster", palette = "Set1") + 
  labs(x = "Pseudotime", y = "Expression") + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

#### Ontology

Test for enrichment of KEGG pathways within each cluster:

```{r}
ens <- rowData(sc1)$gene_id[match(rownames(sp1), rownames(sc1))]

ids <- mapIds(org.Mm.eg.db, keys = ens, column = "ENTREZID", keytype = "ENSEMBL")

ids <- split(ids, sp1$cluster)

res <- lapply(ids, function(x) calculateKEGG(query = x, subject = uni))

plt <- list(
  visualiseKEGG(res$`1`, n = 10) + ggtitle("Cluster 1 - KEGG Pathway"),
  visualiseKEGG(res$`2`, n = 10) + ggtitle("Cluster 2 - KEGG Pathway"),
  visualiseKEGG(res$`3`, n = 10) + ggtitle("Cluster 3 - KEGG Pathway")
)

patchwork::wrap_plots(plt, ncol = 1)
```

### Lineage 2

#### Clustering

Extract association test results from experiment metadata:

```{r}
res <- metadata(sce)$associationTest
```

Create data frame of association test results for slingPseudotime_2 lineage:

```{r}
sp2 <- data.frame(
  waldStat = res$slingPseudotime_2$waldStat_2, 
  df = res$slingPseudotime_2$df_2, 
  pvalue = res$slingPseudotime_2$pvalue_2, 
  fdr = p.adjust(res$slingPseudotime_2$pvalue_2, method = "fdr"),
  row.names = rownames(res$slingPseudotime_2)
)
```

Reorder experiment data by slingPseudotime_2 lineage:

```{r}
sc2 <- reorder(sce, subset = rownames(sp2), select = sce$slingPseudotime_2)

use <- !is.na(sc2$slingPseudotime_2)

sc2 <- sc2[, use]
```

Cluster expression patterns of significant genes in slingPseudotime_2 lineage:

```{r}
set.seed(525499208)

km2 <- kmeans(x = assay(sc2, "rescaled"), centers = 3)

sp2$cluster <- km2$cluster[rownames(sp2)]
```

Plot clustered expression patterns to see how they behave across the slingPseudotime_2 lineage:

```{r}
clust.centroid <- function(i, assay, cluster) {
  ind <- (cluster == i)
  colMeans(assay[ind, ])
}

lvl <- unique(km2$cluster)

df2 <- sapply(lvl, clust.centroid, assay(sc2, "rescaled"), km2$cluster)

df2 <- melt(df2)

colnames(df2) <- c("cell", "cluster", "exprs")

df2$pseudotime <- sc2$slingPseudotime_2[df2$cell]

ggplot(df2, aes(pseudotime, exprs, group = cluster, colour = as.factor(cluster))) + 
  geom_point() + 
  scale_colour_brewer(name = "Cluster", palette = "Set1") + 
  labs(x = "Pseudotime", y = "Expression") + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

#### Ontology

Test for enrichment of KEGG pathways within each cluster:

```{r}
ens <- rowData(sc2)$gene_id[match(rownames(sp2), rownames(sc2))]

ids <- mapIds(org.Mm.eg.db, keys = ens, column = "ENTREZID", keytype = "ENSEMBL")

ids <- split(ids, sp2$cluster)

res <- lapply(ids, function(x) calculateKEGG(query = x, subject = uni))

plt <- list(
  visualiseKEGG(res$`1`, n = 10) + ggtitle("Cluster 1 - KEGG Pathway"),
  visualiseKEGG(res$`2`, n = 10) + ggtitle("Cluster 2 - KEGG Pathway"),
  visualiseKEGG(res$`3`, n = 10) + ggtitle("Cluster 3 - KEGG Pathway")
)

patchwork::wrap_plots(plt, ncol = 1)
```
